---
title: "Perform DE analysis"
subtitle: "Compute individual DEs using Limma and then perform meta-analysis via Linear (Mixed-Effects) Models (RMA)"
author: "Shahin Mohammadi"
date: "Run on `r Sys.time()`"
documentclass: article
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
library(ACTIONet)
library(data.table)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(corrplot)
library(limma)
library(muscat)
library(metafor)
library(ggcorrplot)
library(synapser)
library(synExtra)
synLogin(rememberMe = TRUE)
source("functions.R")
})

```


# Setup environment
```{r}
dataset.path = "~/submission/datasets/"
figures.path = "~/submission/figures"
tables.path = "~/submission/tables"
input.path = "~/submission/input"

# Thresholds on the FDR-corrected meta-analysis results
min.capture.rate = 0.05 # only keep genes with at least 5% capture rate across all cells

# DE gene threshold for meta-anlaysis
pval_threshold = 0.05
logFC_threshold = 0.1

# DE formula
form = as.formula("~ Phenotype + Age.group + Gender + PMI.z + SampleQuality") # SampleQuality/Depth ~ scale(mean_umis)

```

```{r}
# Load pseudobulk samples
pb.sce = loadDataset("pseudobulk_mean_logcounts_filtered", dataset.path = dataset.path)
ACTIONet_summary = loadDataset("ACTIONet_summary", dataset.path = dataset.path)

celltypes.df = loadDataset("celltype_stats", dataset.path = file.path(tables.path, "rds"))
celltype_colors = celltypes.df$color
names(celltype_colors) = celltypes.df$celltype

```


# Compute DE of individual datasets
```{r}
  pb.sce.filtered = pb.sce[rowData(pb.sce)$capture.rate > min.capture.rate, ]

  resDE = lapply( levels(pb.sce.filtered$Cohort), function(chrt){
  	keep.ids = colnames(pb.sce.filtered)[pb.sce.filtered$Cohort == chrt]
  	pb.sce.filtered_sub = pb.sce.filtered[,keep.ids]
    sample.metadata = droplevels(data.frame(colData(pb.sce.filtered_sub)))
  	
    sample.metadata$SampleQuality = scale(sample.metadata$umis) # pb.sce.filtered_sub$
    sample.metadata$PMI.z = scale(sample.metadata$PMI)
    ll = c("Young", "Old")
    groups<- factor(ll[as.numeric(cut(sample.metadata$Age,breaks = 2))], ll)
    sample.metadata$Age.group = groups

  	design.mat <- model.matrix(form, data = sample.metadata)
  	colnames(design.mat)[1] = c("Intercept")
  	contrast.mat <- makeContrasts(contrasts = "PhenotypeSZ", levels = design.mat)
  
    corrected_assays = lapply(1:length(assays(pb.sce.filtered_sub)), function(k) {
      print(k)
      X = assays(pb.sce.filtered_sub)[[k]]
      Xres = limma::removeBatchEffect(X, batch = sample.metadata$Batch, batch2 = sample.metadata$HTO, design = model.matrix(~0. + sample.metadata$Phenotype))
      rownames(Xres) = rownames(pb.sce.filtered_sub)
      colnames(Xres) = colnames(pb.sce.filtered_sub)
      return(Xres)
    })
    names(corrected_assays) = names(assays(pb.sce.filtered_sub))
    assays(pb.sce.filtered_sub) = corrected_assays

  
  	df = pbDS(pb.sce.filtered_sub, method = "limma-trend", min_cells = 10, design = design.mat, contrast =  contrast.mat, filter = "both")
  	
  	return(df)
  })
  names(resDE) = levels(colData(pb.sce.filtered)$Cohort)

```



# Export full DE analysis per cohort 
```{r}
storeDataset(resDE, name = "Cohort_specific_DE_results_final", dataset.path = dataset.path)

```

```{r}
resDE.old = readr::read_rds("~/results/datasets/Cohort_specific_DE_results_final.rds")

```


# Export unfiltered cohort-specific tables
```{r}
for(ds in 1:length(resDE)) {
  print(names(resDE)[[ds]])
  
  Up.DFs = lapply(1:length(resDE[[ds]]$table$PhenotypeSZ), function(i) {
    res = resDE[[ds]]$table$PhenotypeSZ[[i]]
    res = res[res$logFC > 0, ]
    res = cbind(data.frame(Gene = rownames(res)), res)
    res = res[order(res$t, decreasing = T), ]
    return(res)  
  })
  names(Up.DFs) = names(resDE[[ds]]$table$PhenotypeSZ)
  storeTable(Up.DFs, name = sprintf("DE_genes_up_%s_complete_set_final", names(resDE)[[ds]]), tables.path = tables.path)

  Down.DFs = lapply(1:length(resDE[[ds]]$table$PhenotypeSZ), function(i) {
    res = resDE[[ds]]$table$PhenotypeSZ[[i]]
    res = res[res$logFC < 0, ]
    res = cbind(data.frame(Gene = rownames(res)), res)
    res = res[order(res$t, decreasing = F), ]
  })
  names(Down.DFs) = names(resDE[[ds]]$table$PhenotypeSZ)
  storeTable(Down.DFs, name = sprintf("DE_genes_down_%s_complete_set_final", names(resDE)[[ds]]), tables.path = tables.path)
}

```


# Prefiltering individual DE results before combining them
```{r}
common.celltypes = intersect(names(resDE$McLean$table$PhenotypeSZ), names(resDE$MtSinai$table$PhenotypeSZ))

filtered.tables = lapply(common.celltypes, function(celltype) {
  tbl1 = resDE[[1]]$table$PhenotypeSZ[[celltype]]
  tbl2 = resDE[[2]]$table$PhenotypeSZ[[celltype]]
  
  genes = intersect(tbl1$gene, tbl2$gene)

  tbl1 = tbl1[match(genes, tbl1$gene), ]
  tbl2 = tbl2[match(genes, tbl2$gene), ]
  
  tbls = list(McClean = tbl1, MtSinai = tbl2) 
  tbls = lapply( tbls, function(tab){
    tab$se = tab$logFC / tab$t
    tab
  })
  
  return(tbls)
})
names(filtered.tables) = common.celltypes

```

# Perform meta-analysis via Linear (Mixed-Effects) Models (RMA)
```{r}
combined.analysis.tables = lapply(names(filtered.tables), function(celltype) {
  print(celltype)
  tbls = filtered.tables[[celltype]]
  
  gene.tbls = lapply(1:nrow(tbls[[1]]), function(i) {
    dfs = lapply(1:length(tbls), function(k) tbls[[k]][i, ])
    df = do.call("rbind", dfs)
  })
  names(gene.tbls) = tbls[[1]]$gene
    
  combined.analysis.tbl = do.call(rbind, lapply(names(gene.tbls), function(gene){
    x = suppressWarnings(metafor::rma(yi=logFC, sei=se, data = gene.tbls[[gene]], method="FE"))
    combined.tbl = data.frame( gene = gene, 
        logFC     = x$beta,
        se        = x$se,
        tstat = x$zval,
        P.Value   = x$pval)
    return(combined.tbl)
  }))
  rownames(combined.analysis.tbl) = names(gene.tbls)
  
  combined.analysis.tbl = combined.analysis.tbl[order(combined.analysis.tbl$P.Value), ]
  
  return(combined.analysis.tbl)
})
names(combined.analysis.tables) = names(filtered.tables)

combined.analysis.tables.v2 = lapply(combined.analysis.tables, function(DF) {
  DF$adj.P.Val = p.adjust(DF$P.Value, method = "fdr")
  return(DF)
})


DF.total = do.call(rbind, combined.analysis.tables.v2)
DF.total$adj.P.Val.global = p.adjust(DF.total$P.Value, method = "fdr")
ff = factor(unlist(lapply(names(combined.analysis.tables.v2), function(celltype) rep(celltype, nrow(combined.analysis.tables.v2[[celltype]])))), names(filtered.tables))
combined.analysis.tables.v3 = split(DF.total, ff)

print(sort(sapply(combined.analysis.tables.v3, function(DF) sum( (DF$adj.P.Val < 0.05) & (abs(DF$logFC) > 0.1) )), decreasing = T)[1:15])
print(sort(sapply(combined.analysis.tables.v3, function(DF) sum( (DF$adj.P.Val.global < 0.05) & (abs(DF$logFC) > 0.1))), decreasing = T)[1:15])

```

# Export meta-analysis results
```{r}
storeDataset(combined.analysis.tables.v3, "meta_analysis_results_final", dataset.path = dataset.path)


```



# Export final tables
```{r}
  Up.DFs = lapply(1:length(combined.analysis.tables.v3), function(i) {
    res = combined.analysis.tables.v3[[i]]
    res = res[(res$logFC > logFC_threshold) & (res$P.Value <= pval_threshold), ]
    res = res[order(res$t, decreasing = T), ]
    res$isSig = res$adj.P.Val <= pval_threshold
    res$isSig.global = res$adj.P.Val.global <= pval_threshold
  return(res)
  })
  names(Up.DFs) = names(combined.analysis.tables.v3)
  storeTable(Up.DFs, name = "DE_genes_up_combined_final", tables.path = tables.path)

  Down.DFs = lapply(1:length(combined.analysis.tables.v3), function(i) {
    res = combined.analysis.tables.v3[[i]]
    res = res[(res$logFC < -logFC_threshold) & (res$P.Value <= pval_threshold), ]
    res = res[order(res$t, decreasing = F), ]
    res$isSig = res$adj.P.Val <= pval_threshold
    res$isSig.global = res$adj.P.Val.global <= pval_threshold
    return(res)
  })
  names(Down.DFs) = names(combined.analysis.tables.v3)
  
  storeTable(Down.DFs, name = "DE_genes_down_combined_final", tables.path = tables.path)
  
```


# Useful if we are running only the last part(s) of the analysis
```{r}
combined.analysis.tables.v3 = loadDataset("meta_analysis_results_final", dataset.path = dataset.path)

```


# Summarize and simplify DE results
```{r}
DE.sc = matrix(0, nrow(pb.sce), length(combined.analysis.tables.v3))
tstats = matrix(0, nrow(pb.sce), length(combined.analysis.tables.v3))
logFC = matrix(0, nrow(pb.sce), length(combined.analysis.tables.v3))
logPvals = matrix(0, nrow(pb.sce), length(combined.analysis.tables.v3))
rownames(DE.sc) = rownames(tstats) = rownames(logFC) = rownames(logPvals) = rownames(pb.sce)
colnames(DE.sc) = colnames(tstats) = colnames(logFC) = colnames(logPvals) = names(combined.analysis.tables.v3)

limma_trend_mean.scores = matrix(0, nrow(pb.sce), length(combined.analysis.tables.v3))
Up.genes = vector("list", length(combined.analysis.tables.v3))
Down.genes = vector("list", length(combined.analysis.tables.v3))
rownames(limma_trend_mean.scores) = rownames(pb.sce)
names(Up.genes) = names(Down.genes) = colnames(limma_trend_mean.scores) = names(combined.analysis.tables.v3)
for(i in 1:length(combined.analysis.tables.v3)) {
	print(i)
	
	tbl = combined.analysis.tables.v3[[i]]

	tstats[tbl$gene, names(combined.analysis.tables.v3)[[i]]] = tbl$tstat
	logFC[tbl$gene, names(combined.analysis.tables.v3)[[i]]] = tbl$logFC
	logPvals[tbl$gene, names(combined.analysis.tables.v3)[[i]]] = -log10(tbl$adj.P.Val)
	
	x = -log10(tbl$adj.P.Val) * sign(tbl$logFC)
	x[abs(tbl$logFC) < logFC_threshold] = 0
	DE.sc[tbl$gene, names(combined.analysis.tables.v3)[[i]]] = x
	
}
limma_trend_mean.scores[is.na(limma_trend_mean.scores)] = 0




Up.genes = lapply(combined.analysis.tables.v3, function(combined.analysis.tbl) {
  combined.analysis.tbl$gene[(combined.analysis.tbl$logFC > logFC_threshold) & (combined.analysis.tbl$adj.P.Val < pval_threshold)]
})
Down.genes = lapply(combined.analysis.tables.v3, function(combined.analysis.tbl) {
  combined.analysis.tbl$gene[(combined.analysis.tbl$logFC < -logFC_threshold) & (combined.analysis.tbl$adj.P.Val < pval_threshold)]
})

DE.new = list(DE.sc = DE.sc, tstats = tstats, logFC = logFC, logPvals = logPvals, Up.genes = Up.genes, Down.genes = Down.genes)


```


# Store final results
```{r}
storeDataset(DE.new, "DE_genes_pseudobulk_final", dataset.path = dataset.path)

```




```{r}
celltypes = names(combined.analysis.tables.v3)

combined.DFs = lapply(celltypes, function(celltype) {
  meta.df = combined.analysis.tables.v3[[celltype]][, 1:6]
  McLean.df = filtered.tables[[celltype]]$McClean
  MtSinai.df = filtered.tables[[celltype]]$MtSinai
  
  gg = sort(unique(meta.df$gene))
  
  DF = cbind(meta.df[match(gg, meta.df$gene), ], McLean.df[match(gg, McLean.df$gene), c(3:7)], MtSinai.df[match(gg, MtSinai.df$gene), c(3:7)])
  colnames(DF)[-1] = c(paste0("Meta_", colnames(DF)[2:ncol(meta.df)]), paste0("McLean_", colnames(DF)[(ncol(meta.df)+1):(ncol(meta.df)+5)]), paste0("MtSinai_", colnames(DF)[(ncol(meta.df)+6):(ncol(meta.df)+10)]))  
  
  return(DF)
})
names(combined.DFs)  = celltypes
storeTable(combined.DFs, name = "DE_genes_merged_tables_final", tables.path = tables.path)
storeDataset(combined.DFs, name = "DE_genes_merged_tables_final", dataset.path = dataset.path)



```

# Again, helpful if we just want to keep going from here
```{r}
combined.DFs = loadDataset(name = "DE_genes_merged_tables_final", dataset.path = dataset.path)

```


```{r eval=FALSE, include=FALSE}
tstats = lapply(combined.DFs, function(DF) {
  mat.tstats = cbind(DF$Meta_tstat, DF$McLean_t, DF$MtSinai_t)
})

DE.masks = lapply(combined.DFs, function(DF) {
  mat.pval = cbind(DF$Meta_adj.P.Val, DF$McLean_p_adj.loc, DF$MtSinai_p_adj.loc)
  mat.logFC = cbind(DF$Meta_logFC, DF$McLean_logFC, DF$MtSinai_logFC)
  mask = (mat.pval < pval_threshold) & (abs(mat.logFC) > logFC_threshold)
})


sum(mask[, 3] & mask[, 2]  &mask[, 1])



```


# Analyze consistency of DE genes between individual and meta analyses
## Compute overlap gene counts and their corresponding p-values
```{r}
DE.genesets = lapply(combined.DFs, function(DF) {
  mat.pval = cbind(DF$Meta_adj.P.Val, DF$McLean_p_adj.loc, DF$MtSinai_p_adj.loc)
  mat.logFC = cbind(DF$Meta_logFC, DF$McLean_logFC, DF$MtSinai_logFC)
  mask = (mat.pval < pval_threshold) & (abs(mat.logFC) > logFC_threshold)
  
  gs = apply(mask, 2, function(x) DF$gene[x > 0])
  names(gs) = c("Meta", "McLean", "MtSinai")
  return(gs)
})

overlap.sizes = t(sapply(DE.genesets, function(gs) {
  mm = SuperExactTest::supertest(gs, n = nrow(pb.sce))
  mm$overlap.sizes
}))

overlap.pvals = t(sapply(DE.genesets, function(gs) {
  mm = SuperExactTest::supertest(gs, n = nrow(pb.sce))
  mm$P.value
}))

colnames(overlap.pvals) = colnames(overlap.sizes) = c("MtSinai", "McLean", "McLean&MtSinai", "Meta", "Meta&MtSinai", "Meta&McLean", "Meta&McLean&MtSinai")

overlaps = list(Sizes = as.data.frame(overlap.sizes), Pvals = as.data.frame(overlap.pvals))
overlaps$Sizes = cbind(data.frame(celltype = names(DE.genesets)), overlaps$Sizes)
overlaps$Pvals = cbind(data.frame(celltype = names(DE.genesets)), overlaps$Pvals)

storeTable(overlaps, name = "DE_genes_overlaps_final", tables.path = tables.path)
storeDataset(overlaps, name = "DE_genes_overlaps_final", dataset.path = dataset.path)


Shared.genesets = lapply(names(combined.DFs), function(celltype) {
  DF = combined.DFs[[celltype]]
  mat.pval = cbind(DF$Meta_adj.P.Val, DF$McLean_p_adj.loc, DF$MtSinai_p_adj.loc)
  mat.logFC = cbind(DF$Meta_logFC, DF$McLean_logFC, DF$MtSinai_logFC)
  mask = (mat.pval < pval_threshold) & (abs(mat.logFC) > logFC_threshold)
  
  gs = apply(mask, 2, function(x) DF$gene[x > 0])  
  
  names(gs) = c("Meta", "McLean", "MtSinai")
  cg = Reduce("intersect", gs)
  sub.df = DF[match(cg, DF$gene), ]
  if(nrow(sub.df) > 0)
    sub.df$Celltype = celltype
  return(sub.df)
})
shared.df = do.call(rbind, Shared.genesets)

storeTable(list(Shared_genes = shared.df), name = "DE_shared_genes_final", tables.path = tables.path)
storeDataset(shared.df, name = "DE_shared_genes_final", dataset.path = dataset.path)

```


```{r}
    # We need to recompute this with "distinct" option for visualization
    overlap.sizes = t(sapply(DE.genesets, function(gs) {
      m = make_comb_mat(list_to_matrix(gs), mode = "distinct")
      x = comb_size(m)[c("001", "010", "011", "100", "101", "110", "111")]
    }))
    overlap.sizes[is.na(overlap.sizes)] = 0
    
    overlap.logPvals = -log10(overlaps$Pvals[, -1])
    overlap.logPvals[is.na(overlap.logPvals)] = 0
    overlap.logPvals[overlap.logPvals > 150] = 150

    Pur.Pal = circlize::colorRamp2(seq(0, 150, length.out = 10), c("white", pals::brewer.purples(9)))

    require(eulerr)
    lapply(common.celltypes, function(celltype) {
      x = overlap.sizes[celltype, ]
      names(x) = colnames(overlap.logPvals)
      logPvals = overlap.logPvals[celltype, ]
      
      CPal = Pur.Pal(logPvals)
      names(CPal) = names(x)
      
      pp = plot(eulerr::venn(x), quantities = TRUE, fills = list(fill = CPal))

      fname = sprintf("%s_DE_overlap_venn", celltype)
      storeFigure(pp, fname, extension = "pdf", width = 6, height = 6, figures.path = file.path(figures.path, "reproducibility"))
      return(NULL)
    });    

    lapply(common.celltypes, function(celltype) {
      x = overlap.sizes[celltype, ]
      names(x) = colnames(overlap.logPvals)
      logPvals = overlap.logPvals[celltype, ]
      
      CPal = Pur.Pal(logPvals)
      names(CPal) = names(x)
      
      pp = plot(euler(x), quantities = TRUE, fills = list(fill = CPal))

      fname = sprintf("%s_DE_overlap_euler", celltype)
      storeFigure(pp, fname, extension = "pdf", width = 6, height = 6, figures.path = file.path(figures.path, "reproducibility"))
      return(NULL)
    })
    
    
```



```{r}
dd = as.data.frame(overlaps$Sizes[, c("McLean", "MtSinai", "Meta")])
dd$Celltype = common.celltypes

df = reshape2::melt(dd)
colnames(df) = c("Celltype", "Dataset", "Genes")
require(ggpubr)
df$Celltype = factor(df$Celltype, intersect(celltypes.df$celltype, common.celltypes))
df$Dataset = factor(df$Dataset, c("McLean", "MtSinai", "Meta"))

gg = ggbarplot(df, "Celltype", "Genes",
  fill = "Dataset", color = "Dataset", palette = "Paired",
  position = position_dodge(0.9)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, color = celltype_colors[levels(df$Celltype)]))

storeFigure(gg, "DE_sizes_batplot_meta_vs_indiv", extension = "pdf", width = 10, height = 5, figures.path = figures.path)

```

